Introduction to Structural Causal Models

### 2021-05-21 presentation_video.Rmd
More stuff
presentation

# Introduction (1)

- Machine learning has provided many insights into different problems 
- One issue is the consideration of 'What are we actually predicting'?
- Mainstream tools are build on association-based learning 
- Associations are not enough for high stake settings 
- We want causation and not correlation

# Introduction (2)

Causal assumptions differentiate causal models from association learning methods.

@@

@@

- System of equations 
- Assignment equation ':=' rather than regular equation '=' 
- Nonparametric SEM 
- Functional form rather than using conditional probabilities
- SCM = PO + Graphical Tools

- Exogenous factors are part of the model specification

# SCM vs SEM

**Error terms**

Regression: Omittable outside factor 
SCM/SEM; Latent influential factor that is pivotal for the model specification but not observable

Consists of graph and assignments: 
Baseline:

@@

# Assumptions

- First, there was no sign to express the assignment equation and people used the '=' and one would e.g. write $A=B$.
- Treating an equation as a **algebraic equation** led to confusion because those have no causal information.
- This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations.
- The problem is that the equation is symmetric.
- The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
- This misconception has caused a lot of challenges.^[For more information see @pearl2009causality]
- As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.

@@

@@

# Implications

\footnotesize

 |Method         | CBN |  SCM
|----------------|---------------------------|------------------------|
| Prediction     | $\boldsymbol{\cdot}$ Unstable                                                                                        | $\boldsymbol{\cdot}$ Stable                 
|                | $\boldsymbol{\cdot}$ Volatile to parameter changes                                                                   | $\boldsymbol{\cdot}$ More Natural Specification 
|                | $\boldsymbol{\cdot}$ Re-Estimate entire model                                                                        | $\boldsymbol{\cdot}$ Only estimate $\Delta$ CM
|                |                                                                                                                      | 
| Intervention   | $\boldsymbol{\cdot}$ Costly for Non-Markovian Models                                                                 | $\boldsymbol{\cdot}$ Pot. Cyclic Representation
|                | $\boldsymbol{\cdot}$ Unstable(Nature CP)                                                                             | $\boldsymbol{\cdot}$ Stable
|                | $\boldsymbol{\cdot}$ Only generic estimates($\Delta$ CP)                                                             | $\boldsymbol{\cdot}$ Context specific 
|                |                                                                                                                      | 
|Counterfactuals | $\boldsymbol{\cdot}$ **Impossible**                                                                                  | $\boldsymbol{\cdot}$ Possible
|                | $\boldsymbol{\cdot}$ no information on latent factors($\epsilon$)                                                    | $\boldsymbol{\cdot}$ Inclusion of latent factors

@@

$$C_1:= f_{c_1}(p,q) \rightarrow C_1:=600$$
$$C_2:= f_{c_2}(a,b)$$
$$E:=f_E(C_1,C_2,N_E)$$

@@

$$C_1:= f_{c_1}(p,q) \rightarrow C_1:=f(\pi)$$
$$C_2:= f_{c_2}(a,b)$$
$$E:=f_E(C_1,C_2,N_E)$$

@@

@@

@@

\tiny
$$
\begin{array}{|l|l|l|l|l|l|}
\hline \text { model } & \begin{array}{l}
\text { predict in IID } \\
\text { setting }
\end{array} & \begin{array}{l}
\text { predict under } \\
\text { changing } \\
\text { distributions / } \\
\text { interventions }
\end{array} & \begin{array}{l}
\text { answer } \\
\text { counter- } \\
\text { factual } \\
\text { questions }
\end{array} & \begin{array}{l}
\text { obtain } \\
\text { physical } \\
\text { insight }
\end{array} & \begin{array}{l}
\text { automatically } \\
\text { learn from } \\
\text { data }
\end{array} \\
\hline \begin{array}{l}
\text { mechanistic } \\
\text { model } \\
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & ? \\
\hline \begin{array}{l}
\text { structural } \\
\text { causal model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{Y} ? ? \\
\hline \begin{array}{l}
\text { causal } \\
\text { graphical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{Y} ? \\
\hline \begin{array}{l}
\text { statistical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} & \mathrm{Y} \\
\hline
\end{array}
$$
\normalsize

@@

>>>>

# Introduction

@@

Table: [@pearl2012causal]

@@

- System of equations with functions
- Assignment equation ':=' rather than regular equation '=' 
- Nonparametric SEM 
- This algebraic equation would imply that $B=A$ because the order has no concrete meaning in algebraic equations (symmetry)
- The initial $'='$ sign was replaced with the ':=' which is asymmetric (Pearl, 2009) and called an **assignment**.
- As mentioned, we define variables as functions e.g. $A=f_A(B,U_A)$.

**Error:* **
Regression: Omittable outside factor 
SCM/SEM; Latent influential factor that is pivotal for the model specification but not observable

@@

@@

Independence:

- noise terms independent
- mechanisms independent (other variables invariant)(local changes)
- change in distribution stems from change in mechanism

- Causal Markov Condition allows to factorize 
- Each Conditional is a causal mechanism

Source: [@peters_elements_2017]

# Assumptions in Causality (2)

@@

@@

$$C_1:= f_{c_1}(p,N_1) \rightarrow C_1:=600$$
$$C_2:= f_{c_2}(a,N_2)$$
$$E:=f_E(C_1,C_2,N_E)$$

@@

$$C_1:= f_{c_1}(p,N_1) \rightarrow C_1:=f(\pi)$$
$$C_2:= f_{c_2}(a,N_2)$$
$$E:=f_E(C_1,C_2,N_E)$$

@@

- missing data problem in PO framework

@@

# Causal Modelling with Differential Equations

Intial Value:

$$
\mathbf{x}\left(t_{0}\right)=\mathbf{x}_{0}
$$

$$
\frac{d \mathbf{x}}{d t}=f(\mathbf{x}), \mathbf{x} \in \mathbb{R}^{d}
$$

$$
\mathbf{x}(t+d t)=\mathbf{x}(t)+d t \cdot f(\mathbf{x}(t))
$$

@@

\scriptsize

$$
\begin{array}{|l|l|l|l|l|}
\hline \text { model } & \begin{array}{l}
\text { predict in IID } \\
\text { setting }
\end{array} & \begin{array}{l}
\text { predict under } \\
\text { changing } \\
\text { distributions / } \\
\text { interventions }
\end{array} & \begin{array}{c}
\text { answer } \\
\text { counter- } \\
\text { factual } \\
\text { questions }
\end{array} & \begin{array}{l}
\text { obtain } \\
\text { physical } \\
\text { insight }
\end{array} \\
\hline \begin{array}{l}
\text { mechanistic } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} \\
\hline \begin{array}{l}
\text { structural } \\
\text { causal model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} \\
\hline \begin{array}{l}
\text { causal } \\
\text { graphical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} \\
\hline \begin{array}{l}
\text { statistical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} \\
\hline
\end{array}
$$
Source: [@peters_elements_2017]
\normalsize

# Extensions

- Mediation Analysis 
- PO-Framework
- Causal Algorithms
- Causal Constraints Model

# Conclusions

@@

# Appendix
